Estimating duration using survival analysis

### 2024-03-20 cis-perfect-testing.tex
Perf test figures

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

Instead, I propose a weakly informative (or vague) prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
The amount of information in a Beta distribution is related to the sum of its parameters.
Here, the sum equals 2, the same as the flat prior case.
The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
The central estimate, of 0.05, is in line with previous estimates that the median duration i in the range 15--20 days~\autocite{cevikShedding}.
This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/vague-prior}
  \caption[Vague prior for the hazard]{The density of Beta(0.1, 1.9) vague prior (left) and the prior predictive survival time (right). The prior predictive survival time is very vague using a Beta(0.1, 1.9) prior. Note the different x-axis to \cref{perf-test:fig:flat-prior}. \label{perf-test:fig:vague-prior}}
\end{figure}

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The prior predictive distribution (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
Using a multivariate normal, as opposed to multiple univariate distribution preserves the correlation between the hazards.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
To estimate the parameters of the multivariate normal, I use the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  \vspace{-2.5cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for the first 15 hazards. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible). ERR... I'M NOT DOING THIS ANYMORE OH NO!!
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. See main text for details of each analysis. \label{perf-test:fig:duration-dist}}
\end{figure}

>>>>

Uninformative Beta priors on the hazard are problematic in this context.
An uninformative prior would be of the form $\lambda_t \dist \BetaDist(\alpha, \alpha)$.
Commonly, $\alpha$ is chosen to be $0.5$ (Jeffreys' Prior) or $1$ (a uniform prior).
Even though these priors are uninformative on the hazard, they become highly informative on the survival. 
Specifically, they tend to favour shorter survival times (see \cref{perf-test:fig:flat-prior}).
An intuitive explanation of this follows from the fact that priors of this form have an expected value of 0.5.
The expected survival time, $\E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right)$, is therefore equal to $0.5^{t-1}$, a quantity that declines rapidly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption[Uninformative priors for the hazard]{
    (A) Density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors.
    (B) Prior predictive survival time (mean and 95\% CrI).
    The implied prior on the survival time is very short when using either prior on the hazard at each time.
  }
 \label{perf-test:fig:flat-prior}
\end{figure}

Instead, I propose a weakly informative (or vague) prior Beta(0.1, 1.9), which has mean 0.05 and minimal information.
The amount of information in a Beta distribution is related to the sum of its parameters.
Here, the sum equals 2, the same as the flat prior case.
The central 95\% probability mass of Beta(0.1, 1.9) is 0.00--0.47.
The central estimate, of 0.05, is in line with previous estimates that the median duration i in the range 15--20 days~\autocite{cevikShedding}.
This prior gives a very vague prior predictive distribution on the survival time (see \cref{perf-test:fig:vague-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/vague-prior}
  \caption[Vague prior for the hazard]{
    (A) Density of Beta(0.1, 1.9) vague prior.
    (B) Prior predictive survival time (mean and 95\% CrI).
    Note the different x-axis to \cref{perf-test:fig:flat-prior}.
  }
  \label{perf-test:fig:vague-prior}
\end{figure}

@@

RW2 prior encodes that the hazard should be linearly changing with some random changes at each time step.
Specifically, the difference on the logit scale between $\lambda_t$ to $\lambda_{t+1}$ is the difference between $\lambda_{t-1}$ and $\lambda_t$ plus some normally distributed noise.
Formally:
\begin{align}
  \logit\lambda_{t+1}
  &= \logit\lambda_t + (\logit\lambda_t - \logit\lambda_{t-1}) + \sigma \epsilon_t &\text{for $t \geq 2$} \\
  &= 2\logit\lambda_t - \logit\lambda_{t-1} + \sigma \epsilon_t \\
  \epsilon_t &\dist \N(0, 1) &\text{for $t \geq 2$}  \\
  \logit\lambda_2 &= \logit\lambda_1 + \epsilon_1 \\\
  \epsilon_1 &\dist \N(\mu_{\epsilon_1}, \sigma_{\epsilon_1}^2) \\
  \logit \lambda_1 &\dist \N(\mu_{\lambda_1}, \sigma_{\lambda_1}^2) \\
  \sigma &\dist \Exponential(1/\mu_\sigma).
\end{align}
The hyperparameters $\mu_{\lambda_1}$ and $\sigma_{\lambda_1}$ specify the prior on the hazard at time 1; $\mu_{\epsilon_1}$ and $\sigma_{\epsilon_1}$ the prior on the initial gradient; and $\mu_\sigma$ the smoothness of the random walk.
I use $\mu_{\lambda_1} = -17.5$, $\sigma_{\lambda_1} = 6$, $\mu_\epsilon = 1.09$, $\sigma_\epsilon = 0.03$, and $\mu_\epsilon = 0.1$ based on the analysis in \cref{E-ATACCC} (see \cref{perf-test:fig:rw2-prior}).
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/rw2-prior}
  \caption[RW2 prior for the hazard]{The prior predictive survival time (mean and 95\% central interval) for the RW2 prior specified in the main text. The posterior mean estimate using ATACCC data in \cref{E-ATACCC} are shown for comparison. \label{perf-test:fig:rw2-prior}}
\end{figure}

@@

I first approximate the \cref{E-ATACCC} posterior estimate of the hazard as a multivariate normal on the logit scale.
This can be viewed as an approximation of Markov melding~\autocite{goudieJoining}.
Using a multivariate normal, as opposed to multiple univariate distribution preserves the correlation between the hazards.
The logit transformation maps from the $[0, 1]$ interval to the full real line.
To estimate the parameters of the multivariate normal, I use the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC-hazard} and \cref{perf-test:fig:approximate-ATACCC-survival}).
\begin{figure}
  \thisfloatpagestyle{empty}
  % \vspace{-1cm}
  \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{cis-perfect-testing/ataccc-approximation-hazard}}
  \caption[Approximating the ATACCC posterior hazard as a logit-normal]{Comparison of the posterior estimate of the hazard from \cref{E-ATACCC} and its approximation with a logit-normal distribution (kernel density estimate smoothed) for hazards between day 4 and 15. The density for smaller hazards is concentrated very close to 0. \label{perf-test:fig:approximate-ATACCC-hazard}}
\end{figure}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/ataccc-approximation-survival}
  \caption[Approximating the ATACCC posterior survival as a logit-normal]{Comparison of the posterior estimate of the survival from \cref{E-ATACCC} and its approximation with a logit-normal distribution (median and central 95\% interval). The survival is a function of all the hazards up until that point, therefore the correlation between the hazards must be well-approximated to have these agree. \label{perf-test:fig:approximate-ATACCC-survival}}
\end{figure}

@@

To form the duration distribution used in the simulation, we combine the two estimates.
The pdf over the first 30 days is proportional to the ATACCC estimate, with the rest proportional to this CIS-based estimates.
% The CIS-based estimates are shifted 3 days in order to make this smooth; this counteracts the above bias of missing the start of the infection (for times above 30 days, the other bias, missing short infections, is negligible). ERR... I'M NOT DOING THIS ANYMORE OH NO!!
Denote by $f_A(t)$ the distribution function estimated in \cref{E-ATACCC} and $f_C(t)$ that from the CIS-based estimates just derived.
Then define:
$$
f_S'(t) = \begin{cases}
	f_A(t) &t \leq 30 \\
	f_C(t) &t > 30
\end{cases}
$$
Then the distribution used in the simulation is the normalised version of this: $f_S(t) = f'_S(t)/\sum_i f_S'(i)$.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/input-duration-dists}
  \caption[Comparison of duration distributions]{Comparison of the duration distribution used for the simulation. ATACCC is the posterior mean from \cref{E-ATACCC}'s analysis, CIS is Sarah Walker's analysis, and Combined is the combination of these used for the simulation. \label{perf-test:fig:duration-dist}}
\end{figure}
